Automatic Parallelization and Transparent Fault Tolerance\\ (\emph{New project} paper)

### 2016-04-08 pubs/tfp16/paper.tex
figure number didn't stick
this is what's getting submitted

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
%
% mkd for preliminary paper notes
\usepackage{enumitem} 
\usepackage{color} % for parenthetical notes while paper is in progress
\usepackage{comment} % comment environment
\usepackage{hyperref} 
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Automatic Parallelization and Transparent Fault Tolerance\\
       (New project paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All project contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\begin{abstract}
A claimed trend is an increasing appreciation and practice of functional
programming in scientific computing, and in particular strict (by default)
functional programming, in various guises.  A second claimed trend in both
scientific computing and functional programming is transparent fault
tolerance.  Our project goal is the demonstration of a light-weight,
higher-order, polymorphic, pure functional language implementation in which we
can experiment with automatic parallelization strategies, varying degrees of
strictness, and mechanisms for transparent fault tolerance.
%We do not consider speculative evaluation, or semantic strictness inferred by
%program analysis, so potential parallelism is dictated by the degree of
%strictness.
\\
\emph{Draft paper notes:}
\begin{enumerate}[before=\itshape]
\item We are reporting work in progress.  At the time of writing we have a
  complete serial implementation, a primitive proof-of-concept parallel
  implementation, and are currently developing a more realistic parallel
  runtime.  We expect to have preliminary results from our new parallel
  implementation by the time of the symposium.  The fault-tolerance mechanisms
  remain future work.

\item While this does not qualify as a student paper, students
  (undergraduate!)  have contributed considerably to this effort, and to the
  writing of this paper, and will continue to do so.  If this paper is
  accepted we plan to have one of these students present it at TiFP.

\item Hopefully in the spirit of a student-friendly symposium, for this draft
  version we have chosen a more informal, conversational style of presentation
  than necessarily appropriate for a final paper, have highlighted student
  involvement and its influence on our approach, and pointed out student
  contributions in particular.
\end{enumerate}

@@

\subsection{A trend: strict but still pure functional programming in HPC}
Interest in functional programming for HPC in general is not new as evidenced
by the Workshop on Functional High Performance Computing, held in conjunction
with the International Conference on Functional Programming, now in its fifth
consecutive year.  A more precise characterization of the trend is that there
is a small but growing understanding by practicing computational scientists
that \emph{pure functional semantics}, in some form, is essential to reining
in the complexity of ever-evolving scientific codes and informing the
designs of new ones.  Such semantics are expressed in a number of ways as
highlighted following.

@@

More generally, \emph{task-based} parallelism is becoming increasingly popular
for HPC, often in conjunction with a higher-level (typically inter-node) bulk
synchronous model usually implemented with the Message Passing Interface
(MPI)~\cite{MPI}\@.  In the imperative world, example systems include the
aforementioned Legion, Intel's Threading Building Blocks~\cite{Reinders:2007},
and the HPX C++11/14 extension~\cite{HPX}.  While the concept of task is
somewhat vague, the general idea is a unit of computation with no or
minimal/constrained side effects.  A function in a pure functional language
could be regarded as the essence of the concept of a task.  All of these
systems are of course in languages with strict function-call semantics.

\subsection{Another trend: transparent fault tolerance}
In the quest for computing platforms approaching exascale, it is widely
recognized that new mechanisms for fault tolerance will need to be built into
the software stack.  Current practice in HPC is checkpoint/restart, wherein at
various points in time sufficient program state is dumped to backing store
such that the program, in case of a fault, can be (manually) restarted from
the last stored state.  Most implementations are explicit, though there has
been some success with transparent (invisible to the programmer) systems.
Wiring checkpoint/restart into a large-scale application is not only
burdensome and error-prone, it can severely warp the engineering of the
application because it must be designed such that consistent, essential state
can be readily captured.  Furthermore, there is a concern that the scalability
of checkpoint/restart is reaching its end as data transfer to backing store
becomes an increasing bottleneck.

@@

We claim, then, that there is a trend, in HPC at least, towards
\emph{strict-by-default pure functional programming}, whether as imposed by
programmer discipline, a parallel runtime system, or the language itself
(e.g.\ strict Haskell).  Second, we claim that there is a trend in
high-performance computing, and by transitivity (in addition to Stewart's
work) in functional programming, towards transparent fault tolerance.

@@

\section{Project and Goals}
Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with varying degrees of
language strictness, and secondarily, with mechanisms for transparent fault
tolerance.  Light-weight has several implications.  First and foremost it
means that \emph{we are not attempting to compete with GHC in any
  respect}---succumbing to the temptation of recreating GHC language, type
system, and runtime features, or significant code optimization, would simply
divert us from our central goals.\footnote{We estimate that we will have, to
  within an order of magnitude, one person-hour of total available development
  time for each person-year that GHC currently represents, or put another way,
  one second for each hour, respectively.}  For another, it means that we are
not undertaking significant morphing of GHC itself: because this is in large
part a student project it must be kept as simple as reasonably possible in
various respects.  Our initial requirements are that
%
\begin{itemize}
\item The implementation must be feasible in terms of available effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling must be properly and consistently implemented;
\item The generated code is C.
\end{itemize}
%
For feasibility, we restrict ourselves to shared-memory implicit parallelism.
The requirement for tail calling is not primarily for efficiency, rather it is
for a reasonable parallel implementation.  Achieving this, and thereby
sufficiently limiting the use of the native C stack, will be discussed at some
length.

@@

Our incarnation of STG, as shown in Figure~\ref{fig:STGsyntax}, is meant to
evoke both an abstract and concrete syntax.  This presentation is adapted from
Marlow and Peyton Jones' \emph{Eval/Apply} paper~\cite{Marlow:2004}, to which
we refer so many times that we will subsequently refer to it as simply
\emph{EA}.  Black hole (\emph{BLACKHOLE}), signifying a \emph{THUNK} being
evaluated, is only a runtime entity and not actually part of our STG language.

The semantics of STG is what one would expect for a higher-order, polymorphic,
pure functional language, with a few exceptions for a Haskell programmer
familiar with unboxed types.  First, \texttt{case} is strict in the scrutinee,
so \texttt{case} is similar to \texttt{pseq} (as opposed to \texttt{seq}) in
Haskell, that is, the scrutinee is evaluated first.  Second, constructors are
not functions, rather they are named in the construction (heap allocation) of
a data object.  More generally, the operational notion of allocating a heap
object is explicit: this is the operational semantics of \texttt{let} bindings.
%As such a variable of unboxed type cannot be let-bound; instead it can be
%bound with \texttt{case}, e.g., \texttt{case e of x\# \{ \_ -> e \}}
%where the \texttt{\#} suffix is merely a convention to signify that
%\texttt{x\#} has unboxed type.  
Finally, deferred evaluation is explicit:
a deferred expression evaluation is encoded by a \emph{THUNK}.

\subsection{Degrees of strictness}
First some words about terminology.  For all degrees of strictness, sharing
(as implied by lazy but not by non-strict) of heap-allocated objects is
preserved so we will not use the term \emph{lazy} and simply refer to
strictness properties.  As is common we will use strict in an operational
sense, meaning that argument(s) are evaluated before being passed in a
function call, rather than the denotational sense, unless stated otherwise.
Similarly, when referring to language semantics we refer by default to the
operational rather than denotational semantics.

@@

In STG the arity of a function is syntactically defined as the number of
manifest formal parameters in a function definition, thus \texttt{id x = x}
has by definition arity one regardless of the type of the argument (e.g., some
function type) it might be applied to.  We distinguish \emph{application},
e.g., \texttt{f x y}, which is a syntactic notion, from actually calling a
function. Thus \texttt{id id x} is an application of \texttt{id} to two
arguments, but the underlying \emph{FUN} (user- or system- defined function)
when actually called is given exactly one argument (the first one).

@@

Given application \texttt{f x y} when might \texttt{x} and \texttt{y} be
evaluated?  In the non-strict semantics evaluation is on demand, after
\texttt{f} has been evaluated, the underlying \emph{FUN} actually called
(i.e., not just partially applied and wrapped up in a \emph{PAP}), and one or
both of those arguments possibly forced in subsequent evaluation (assuming
that they weren't forced in the evaluation of \texttt{f} itself).

In a strict semantics one could devise various evaluation orders, but as far
as distinguishing termination properties it seems these can be distilled down
to two reasonable possibilities, namely whether on construction of a
\emph{PAP} the (new) arguments are forced or not.  In other words, the
question is whether application is strict, or actual function call is strict.
These are in fact distinguishable because (in Haskell) the first argument to
\texttt{seq} might evaluate to a \emph{PAP}.

This then gives three reasonable operational modes which are strictly ordered
in terms of termination properties: non-strict; function strict, in which only
those arguments needed by a function call are evaluated; and application
strict, in which all arguments of an application are evaluated.  The other
dimension we vary is whether or not constructors are strict.  Other dimensions
are possible, for example whether \texttt{let} bindings are strict, but we
restrict ourselves to function/application and constructor strictness.  The
STG machinery makes the choice of evaluation strategy easy to change, though
changing code generation can be more efficient by avoiding the unnecessary
creation of \emph{THUNKS} as discussed in Appendix~A.

@@

\subsubsection{Proper tail calls in C.}
Parallelism is achieved by multi-threading; our model is based on Harris et
al.'s.~\cite{Harris:2005}; later developments by Marlow et al.\ are beyond our
aspirations~\cite{Marlow:2009,Marlow:2011}.  To achieve efficient parallel
execution they identified the need for the ability to examine and unwind the
stack in quashing concurrent evaluation of the same \emph{THUNK}.  This
requires that use of the C stack be highly constrained if we are to avoid
complex and ugly hacks using, e.g., \emph{setjmp/longjmp}, or non-portable
code to directly manipulate the C stack, complicating an already non-trivial
implementation effort.  We therefore must limit the use of the C stack to
calls to the runtime, in other words, when executing code generated from the
user program (and not in a runtime call), there must be no growing of (pushing
return addresses or data onto) the C stack.
%
To clearly distinguish between a notional C call (a syntactic concept) and a
machine-level call (pushing a return address onto the native C stack) we will
refer to the latter as \emph{call-with-return}, and to a machine-level tail
call as \emph{jump}.

@@

\emph{EA} describes a family of generic C-{}- \emph{stgApply} runtime
functions used for applications of non-known (not statically identifiable)
functions, and applications of known functions to other than their arity
number of arguments.  Here the object to be applied is first evaluated with
call-with-return.  When a resulting \emph{FUN} is applied to more than its
arity number of arguments, it is called-with-return with its arity number of
arguments, and the resulting value applied to the remaining arguments.  When
considering more strict semantics call-with-return is even more convenient
because arguments can be evaluated in a tight loop.  Unfortunately these uses
of call-with-return can result in the C stack growing without bound.  Our
solution uses a single C \emph{stgApply} function with no calls-with-return.
This is straightforwardly realizable because the stack is explicit, and with
modern C compiler call optimization, the continuation style enables a proper
\emph{goto} mechanism where the \emph{code labels}---entities that name blocks
of code, can be treated as first class (e.g.\ stored), and serve as targets of
a jump---are C function addresses (c.f.\ \S 6.1~\cite{PJ-stockhardware}).

@@

We are currently developing the parallel runtime.  Parallel garbage collection
is likely out of scope.

@@

\subsubsection{Strict but still pure.}
The concept of strict but still pure in some sense dates back to the concept
of applicative order reduction in the lambda calculus.  In terms of a
notionally concrete programming language the first thorough explications may
be Backus' FP~\cite{Backus:1978} and follow-on FL
implementation~\cite{AikenFL,FLreport89}.
%
A notable argument for the potential merits of a strict-by-default
pure functional language over a non-strict one is by Sheard, who also provided
a very simple prototype implementation~\cite{Sheard:2003}.
%
A plug-in for GHC (prior to 8.0) exists for making functions, but not data
structures, strict using a transformation on GHC Core that recursively expands
non-recursive let bindings into case bindings~\cite{strict-ghc-plugin}.  Thus
it seems that while it makes functions denotationally strict, function
arguments are still evaluated after function call.
%
The Disciplined Disciple Compiler project appears to intend implementation of
a strict, higher order functional language (or family of languages) compiler
that has been underway for some time, but the emphasis appears to be on a much
more complex type system for handling stateful computation, and the prospects
for its eventual completion are unknown~\cite{disciplined-disciple}.
%
NVIDIA's NOVA language, and corresponding compiler, implements a strict
(``call by value''), pure, polymorphic, higher-order functional language with
built-in parallel operators (e.g., \emph{map}, \emph{reduce}, and
\emph{scan})~\cite{NVIDIA:2013}. Three back-ends were implemented: for
sequential C, parallel (multi-threaded) C, and CUDA C, the lattermost
targeting NVIDIA GPUs.

\subsubsection{GHC as a front-end.}
The Intel Labs Haskell Research Compiler (HRC) is the most closely related
work to ours of which we are aware~\cite{Liu:2013,Petersen:2013}.  As in our
planned approach they use GHC as the front end, but exit the GHC toolchain
at the final GHC Core stage, just before transformation to STG.  The essence
of their effort is to target an existing, highly optimizing, Intel functional
language compiler that is ``largely language agnostic'' and performs (among
many other optimization) SIMD vectorization.

@@

As mentioned, a current near-term goal is to use GHC as our front-end to
generate STG (or Core) as input to our system.  In the meantime, as a stop-gap
measure to make the generation of test cases easier, we have implemented a
\emph{mini-Haskell} front-end that supports a subset of standard Haskell
syntax with \emph{MagicHash} enabled, with a single extension.  Recall that
\emph{MagicHash} allows \texttt{\#} to be a suffix of an identifier, or denote
an unboxed literal value.  Our extension (to STG as well) is the provision of
the optional \texttt{unboxed} keyword in datatype definitions to allow
user-defined unboxed data types, e.g.,
\begin{verbatim}
  data unboxed Bool# = False# | True# .
\end{verbatim}

@@

An informal specification of the mini-Haskell grammar is given in
%something really broken
%Figure~\ref{fig:miniHaskell}.  
Figure~2.  
There is significant overlap with STG resulting
from the structural similarities, but the redundancies are included for
completeness.

@@

Because mini-Haskell was designed to supplement STG for test generation, it
makes sense to talk about it in terms of the differences between the two and
how mini-Haskell is translated into STG for the back-end.  Here, the
differences are given in order of perceived utility, or importance to someone
transitioning from STG to mini-Haskell.

@@

\begin{verbatim}
append = FUN(l1 l2 ->
            case l1 of
              { Nil -> l2;
                Cons hd tl -> 
                  let { rec = THUNK(append tl l2);
                        result = CON(Cons hd rec) }
                  in result });
\end{verbatim}

In mini-Haskell, expressions may be applied to other expressions, so the
recursive call to \texttt{append} can be parenthesized. There is no need for the
\texttt{let} expression:

\begin{verbatim}
append l1 l2 = case l1 of
                Nil -> l2
                Cons hd tl -> Cons hd (append tl l2)
\end{verbatim}

This also serves as a good example of how expression to expression application is
transformed into STG.  Any non-atomic expression in an application expression is
bound to a heap object in a generated \texttt{let} expression.  These
expressions are thus properly ``atomized'' for STG.  The above mini-Haskell code
is transformed into STG that is structurally identical to the STG code that
precedes it.

@@

STG supports pattern matching but only in a simple form and only in
\texttt{case} expressions.  mini-Haskell extends this to something approaching
what is found in Haskell.  Functions, including lambda expressions, may match on
all their arguments. Named functions may have multiple definitions to exhaust
the combinations of patterns in their arguments and patterns in all contexts
(functions and case expressions) may be nested arbitrarily deep.  This feature
greatly simplifies the task of defining more complex behavior.  Without it,
each matched parameter of a function requires an additional level of nested
\texttt{case} expressions. The same is true of each level of nesting within a
single pattern.

Even relatively simple functions can be painful to write, such as \texttt{unzip}.

>>>>

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
%
% mkd for preliminary paper notes
\usepackage{enumitem} 
\usepackage{color} % for parenthetical notes while paper is in progress
\usepackage{comment} % comment environment
\usepackage{hyperref} 
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Automatic Parallelization and Transparent Fault Tolerance\\
       (\emph{New project} paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{Ringo, Anderson, and Marks are undergraduate students.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\begin{abstract}
A claimed trend is an increasing appreciation and practice of functional
programming in scientific computing, and in particular strict (by default)
functional programming, in various guises.  
%
%A second claimed trend is, given a strict functional semantics, in the
%development of automatically parallelizing runtime systems.
Besides making programs easier to reason about, this enables automatic
parallelization to various degrees.
%
A second claimed trend in both scientific computing and functional programming
is transparent fault tolerance.  
%
Our project goal is the demonstration of a light-weight, higher-order,
polymorphic, pure functional language implementation in which we can
experiment with automatic parallelization strategies, varying degrees of
strictness, and mechanisms for transparent fault tolerance.
%We do not consider speculative evaluation, or semantic strictness inferred by
%program analysis, so potential parallelism is dictated by the degree of
%strictness.
\\
\emph{Draft paper notes:}
\begin{enumerate}[before=\itshape]
\item We are reporting work in progress.  At the time of writing we have a
  complete serial implementation, a primitive proof-of-concept parallel
  implementation, and are currently developing a more realistic parallel
  runtime.  We expect to have preliminary results from our new parallel
  implementation by the time of the symposium.  The fault-tolerance mechanisms
  remain future work.

\item Hopefully in the spirit of a student-friendly symposium, for this draft
  version we have chosen a more informal, conversational style of presentation
  than necessarily appropriate for a final paper, have highlighted student
  preparation, involvement, and its influence on our approach, and pointed out
  some specific student contributions.

\item While this \emph{does not qualify} as a student paper, students
  (undergraduate!) have contributed considerably to this effort, and to the
  writing of this paper, and will continue to do so.  If this paper is
  accepted we plan to have one of these students present it at TiFP.
\end{enumerate}

@@

\subsection{A trend: strict, pure functional programming in HPC and FP}
Interest in functional programming for HPC in general is not new as evidenced
by the Workshop on Functional High Performance Computing, held in conjunction
with the International Conference on Functional Programming, now in its fifth
consecutive year.  A more precise characterization of the trend is that there
is a small but growing understanding by practicing computational scientists
that \emph{pure functional semantics}, in some form, is essential to reining
in the complexity of ever-evolving scientific codes and informing the designs
of new ones, and that strict semantics tend to be preferred.  Such semantics
are expressed in a number of ways as highlighted following.

@@

More generally, \emph{task-based} parallelism is becoming increasingly popular
for HPC, often in conjunction with a higher-level (typically inter-node) bulk
synchronous model usually implemented with the Message Passing Interface
(MPI)~\cite{MPI}\@.  In the imperative world, example systems include the
aforementioned Legion, Intel's Threading Building Blocks~\cite{Reinders:2007},
and the HPX C++11/14 extension~\cite{HPX}, all of which also seek to
make parallelism automatic and transparent.  While the concept of task is
somewhat vague, the general idea is a unit of computation with no or
minimal/constrained side effects.  A function in a pure functional language
could be regarded as the essence of the concept of a task.  All of these
systems are of course in languages with strict function-call semantics.

\subsection{Another trend: transparent fault tolerance in HPC and FP}
In the quest for computing platforms approaching exascale, it is widely
recognized that new mechanisms for fault tolerance will need to be built into
the software stack.  Current practice in HPC is checkpoint/restart, wherein at
various points in time sufficient program state is dumped to backing store
such that the program, in case of a fault, can be (manually) restarted from
the last stored state.  Most implementations are explicit, though there has
been some success with transparent (invisible to the programmer) systems.
Wiring checkpoint/restart into a large-scale application is not only
burdensome and error-prone, it can severely warp the engineering of the
application because it must be designed such that consistent, essential state
can be readily captured.  Furthermore, there is a concern that the scalability
of checkpoint/restart is reaching its end as data transfer to backing store
becomes an increasing bottleneck.

@@

We claim, then, that there is a trend, in HPC at least, towards
\emph{strict-by-default pure functional programming} and automatic
parallelism.
%whether as imposed by programmer discipline, a parallel runtime system, or the
%language itself (e.g.\ strict Haskell).
Second, we claim that there is a trend in
high-performance computing, and by transitivity (in addition to Stewart's
work) in functional programming, towards transparent fault tolerance.

@@

\section{Project and Goals}
Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with varying degrees of
language strictness, and secondarily, with mechanisms for transparent fault
tolerance.  Light-weight has several implications.  First and foremost it
means that \emph{we are not attempting to compete with GHC in any
  respect}---succumbing to the temptation of recreating GHC language, type
system, and runtime features, or significant code optimization, would simply
divert us from our central goals.\footnote{We estimate that we will have, to
  within an order of magnitude, one person-hour of total available development
  time for each person-year that GHC currently represents, or put another way,
  one second for each hour, respectively.}  For another, it means that we are
not undertaking significant morphing of GHC itself: because this is in large
part an undergraduate student project it must be kept as simple as reasonably
possible.

Our initial requirements are that
%
\begin{itemize}
\item The implementation must be feasible in terms of available effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling must be properly and consistently implemented;
\item The generated code is C.
\end{itemize}
%
For feasibility we restrict ourselves to shared-memory implicit parallelism.
The requirement for tail calling is not primarily for efficiency, rather it is
for a reasonable parallel implementation.  Achieving this, and thereby
sufficiently limiting the use of the native C stack, will be discussed at some
length.

@@

Our incarnation of STG, as shown in Figure~\ref{fig:STGsyntax}, is meant to
evoke both an abstract and concrete syntax.  This formulation and presentation
is adapted from Marlow and Peyton Jones' \emph{Eval/Apply}
paper~\cite{Marlow:2004}, to which we refer so many times that we will
subsequently refer to it as simply \emph{EA}.  A (\emph{BLACKHOLE}),
signifying a \emph{THUNK} being evaluated, is only a runtime entity and
not part of the language.

With a few exceptions the semantics of STG is what a Haskell programmer
familiar with unboxed types would expect for a higher-order, polymorphic, pure
functional language.  Points to note are, first, that \texttt{case} is strict
in the scrutinee (the \emph{e} in \texttt{case} \emph{e} \texttt{of}), so \texttt{case} is
similar to \texttt{pseq} (as opposed to \texttt{seq}) in Haskell, that is, the
scrutinee is evaluated first.  Second, constructors are not functions, rather
they are named in the construction (heap allocation) of a data object.  More
generally, the operational notion of allocating a heap object is explicit:
this is the operational semantics of \texttt{let} bindings.
%As such a variable of unboxed type cannot be let-bound; instead it can be
%bound with \texttt{case}, e.g., \texttt{case e of x\# \{ \_ -> e \}}
%where the \texttt{\#} suffix is merely a convention to signify that
%\texttt{x\#} has unboxed type.  
Finally, deferred evaluation is explicit:
a deferred expression evaluation is encoded as a \emph{THUNK}.

\subsection{Degrees of strictness}
First some words about terminology.  For all degrees of strictness, sharing
(as implied by lazy but not by non-strict) of heap-allocated objects is
preserved so we will not use the term lazy and simply refer to
strictness properties.  As is common we will use strict in an operational
sense, meaning that argument(s) are evaluated before being passed in a
function call, rather than the denotational sense, unless stated otherwise.
Similarly, when referring to language semantics we refer by default to the
operational rather than denotational semantics.

@@

In STG the arity of a function is defined as the number of formal parameters
in the function definition, thus \texttt{id x = x} has by definition arity one
regardless of the type of the argument (e.g., some function type) it might be
applied to.  We distinguish \emph{application}, e.g., \texttt{f x y}, which is
a syntactic notion, from actual function \emph{call}. Thus \texttt{id id x} is
an application of \texttt{id} to two arguments, but the underlying \emph{FUN}
(user- or system- defined function) when called is given exactly one
argument (the first one).

@@

Given application \texttt{f x y}, when might \texttt{x} and \texttt{y} be
evaluated?  In the non-strict semantics evaluation is on demand, after
\texttt{f} has been evaluated, the underlying \emph{FUN} actually called
(i.e., not just partially applied and wrapped up in a \emph{PAP}), and one or
both of those arguments possibly forced in subsequent evaluation (assuming
that they weren't forced in the evaluation of \texttt{f} itself).

In a strict semantics one could devise various evaluation orders, but as far
as distinguishing termination properties it seems that these can be distilled
down to two reasonable possibilities, namely whether on construction of a
\emph{PAP} the (new) arguments are forced or not.  In other words, the
question is whether application is strict, or function call is strict.  These
are in fact distinguishable because (in Haskell) the first argument to
\texttt{seq} might evaluate to a \emph{PAP}.

This then gives three reasonable operational modes which are strictly ordered
in terms of termination properties: 
non-strict, function strict, and application strict.
% non-strict; function strict, in which only
%those arguments needed by a function call are evaluated; and application
%strict, in which all arguments of an application are evaluated.  
The other dimension we vary is whether or not constructors are strict.  Other
dimensions are possible, for example whether \texttt{let} bindings are strict,
but we restrict ourselves to function/application and constructor strictness.
The STG machinery makes the choice of evaluation strategy easy to change,
though changing code generation can be more efficient by avoiding the
unnecessary creation of \emph{THUNKS} as discussed in Appendix~A.

@@

\subsubsection{Proper tail calls in C.}
Parallelism is achieved by multi-threading; our model is based on Harris et
al.'s.~\cite{Harris:2005}; later developments by Marlow et al.\ are beyond our
aspirations~\cite{Marlow:2009,Marlow:2011}.  To achieve efficient parallel
execution they identified the need for the ability to examine and unwind the
stack in quashing concurrent evaluation of the same \emph{THUNK}\@.  This
requires that use of the C stack be highly constrained if we are to avoid
complex and ugly hacks using, e.g., \emph{setjmp/longjmp}, or non-portable
code to directly manipulate the C stack, complicating an already non-trivial
implementation effort.  We therefore must limit the use of the C stack to
calls to the runtime, in other words, when executing code generated from the
user program (and not in a runtime call), there must be no growing of (pushing
return addresses or data onto) the C stack.
%
To clearly distinguish between a notional C call (a syntactic concept) and a
machine-level call (pushing a return address onto the native C stack) we will
refer to the latter as \emph{call-with-return}, and to a machine-level tail
call as \emph{jump}.

@@

\emph{EA} describes a family of generic C-{}- \emph{stgApply} runtime
functions used for applications of non-known (not statically identifiable)
functions, and applications of known functions to other than their arity
number of arguments.  There the object to be applied is first evaluated with
call-with-return.  When a resulting \emph{FUN} is applied to more than its
arity number of arguments, it is called-with-return with its arity number of
arguments, and the resulting value applied to the remaining arguments.  When
considering more strict semantics call-with-return is even more convenient
because arguments can be evaluated in a tight loop.  Unfortunately these uses
of call-with-return can result in the C stack growing without bound.  Our
solution uses a single C \emph{stgApply} function with no calls-with-return.
This is straightforwardly realizable because the stack is explicit, and with
modern C compiler call optimization, the continuation style enables a jump
(\emph{go to}) mechanism where the \emph{code labels}---entities that name
blocks of code, can be treated as first class (e.g., stored for later use),
and serve as targets of a jump---are C function addresses (c.f.\ \S
6.1~\cite{PJ-stockhardware}).

@@

We are currently developing the parallel runtime.  Parallel garbage collection
is likely out of scope.
%
The fault-tolerance mechanisms remain future work.

@@

\subsubsection{Strict but still pure.}
The concept of strict but still pure in some sense dates back to the concept
of applicative order reduction in the lambda calculus.  In terms of a
notionally concrete programming language the first thorough explications may
be Backus' FP~\cite{Backus:1978} and follow-on FL
implementation~\cite{AikenFL,FLreport89}.
%
A notable argument for the potential merits of a strict-by-default
pure functional language over a non-strict one is by Sheard, who also provided
a very simple prototype implementation~\cite{Sheard:2003}.
%
A plug-in for GHC (prior to 8.0) exists that makes function bodies strict in
their non-recursive \texttt{let} bindings using a transformation on GHC Core
that recursively expands those \texttt{let} bindings into \texttt{case}
bindings~\cite{Bolingbroke:2008,strict-ghc-plugin}.
%
The Disciplined Disciple Compiler project appears to intend implementation of
a strict, higher order functional language (or family of languages) compiler
that has been underway for some time, but the emphasis appears to be on a much
more complex type system for handling stateful computation, and the prospects
for its eventual completion are unknown~\cite{disciplined-disciple}.

\subsubsection{Strict, pure, auto-parallelizing.}
NVIDIA's NOVA language, and corresponding compiler, implements a strict
(``call by value''), pure, polymorphic, higher-order functional language with
built-in parallel operators (e.g., \emph{map}, \emph{reduce}, and
\emph{scan})~\cite{NVIDIA:2013}. Three back-ends were implemented: for
sequential C, parallel (multi-threaded) C, and CUDA C, the lattermost
targeting NVIDIA GPUs.

\subsubsection{GHC as a front-end.}
The Intel Labs Haskell Research Compiler (HRC) is the most closely related
work to ours of which we are aware~\cite{Liu:2013,Petersen:2013}.  As in our
planned approach they use GHC as the front end, but exit the GHC toolchain
at the final GHC Core stage, just before transformation to STG.  The essence
of their effort is to target an existing, highly optimizing, Intel functional
language compiler that is ``largely language agnostic'' and performs (among
many other optimizations) SIMD vectorization.

@@

As mentioned, a current near-term goal is to use GHC as our front-end to
generate STG (or Core) as input to our system.  In the meantime, as a stop-gap
measure to make the generation of test cases easier, we have implemented a
\emph{mini-Haskell} front-end that supports a subset of standard Haskell
syntax with \emph{MagicHash} enabled, with a single extension.  Recall that
\emph{MagicHash} allows \texttt{\#} to be a suffix of an identifier, or denote
an unboxed literal value.  Our extension (as for STG) is the provision of
the optional \texttt{unboxed} keyword in datatype definitions to allow
user-defined unboxed data types, e.g.,
\begin{verbatim}
  data unboxed Bool# = False# | True# .
\end{verbatim}

@@

An informal specification of the mini-Haskell grammar is given in
Figure~\ref{fig:miniHaskell}.  There is significant overlap with STG resulting
from the structural similarities, but the redundancies are included for
completeness.

@@

Because mini-Haskell was designed to supplement STG for test generation, it
makes sense to talk about it in terms of the differences between the two and
how mini-Haskell is translated into STG for the back-end.  Here, the
differences are given in order of perceived utility, or importance in
transitioning from STG to mini-Haskell.

@@

\begin{verbatim}
append = FUN(l1 l2 ->
            case l1 of
              { Nil -> l2;
                Cons hd tl -> 
                  let { rec = THUNK(append tl l2);
                        result = CON(Cons hd rec) }
                  in result });
\end{verbatim}
In mini-Haskell, expressions may be applied to other expressions, so the
recursive call to \texttt{append} can be parenthesized. There is no need for the
\texttt{let} expression:
\begin{verbatim}
append l1 l2 = case l1 of
                 Nil -> l2
                 Cons hd tl -> Cons hd (append tl l2)
\end{verbatim}
This also serves as a good example of how expression to expression application is
transformed into STG.  Any non-atomic expression in an application expression is
bound to a heap object in a generated \texttt{let} expression.  These
expressions are thus properly ``atomized'' for STG.  The above mini-Haskell code
is transformed into STG that is structurally identical to the STG code that
precedes it.

@@

STG supports pattern matching but only in a simple form and only in
\texttt{case} expressions.  Mini-Haskell extends this to something approaching
what is found in Haskell.  Functions, including lambda expressions, may match on
all their arguments. Named functions may have multiple definitions to exhaust
the combinations of patterns in their arguments and patterns in all contexts
(functions and case expressions) may be nested arbitrarily deep.  This feature
greatly simplifies the task of defining more complex behavior.  Without it,
each matched parameter of a function requires an additional level of nested
\texttt{case} expressions. The same is true of each level of nesting within a
single pattern.
